Circularity in Voxel-Based Analysis

### 2012-01-05 ANTsTBSS/paperRevision.tex
New Safety Suit album---good stuff.

\usepackage{color}
\usepackage{multirow,booktabs,ctable,array}
\usepackage{lscape}
\usepackage{amsmath}
\usepackage{lineno}
\usepackage{ulem}
\usepackage{setspace}
\usepackage{listings}
\usepackage{float}

@@

%% The amssymb package provides various useful mathematical symbols
\usepackage{amssymb}
%% The amsthm package provides extended theorem environments
% \usepackage{amsthm}

 \usepackage{makecell}

@@

\begin{abstract}
Recent discussions within the neuroimaging community have highlighted the
problematic presence of selection bias in experimental design.  Although 
initially centering around the selection of voxels during the course of
fMRI studies, we demonstrate how this bias can 
potentially corrupt voxel-based analysis with a particular emphasis on
population studies of fractional anisotropy (FA).  
For such studies, template-based registration plays a critical role
in which a representative template serves as the normalized space for 
group alignment.  Following 
computation of FA, each subject's FA map is registered to the FA template prior to 
performing statistical comparisons between different groups.
We illustrate how this analysis strategy can induce selection
bias which varies in strength with the similarity metric used in
registration.  Specifically, we analytically demonstrate that the popular sum of squared
difference (SSD) intensity metric explicitly maximizes effect size rather than
anatomical alignment---an experimental design flaw referred to as `circularity bias.'  
Through the use of both simulated diffusion data and real data involving
traumatic brain injury survivors, we explore the magnitude of this bias
and how it varies with similarity metric under the hypothesis that while
SSD-related metrics, such as Demons, will manifest similar effects, 
other metrics which 
are not formulated based on absolute intensity differences 
(e.g. mutual information (MI) and cross correlation (CC)) will produce
less of an effect. 
Consequently, given the variability in voxel-based analysis outcomes 
with similarity metric choice,
we caution researchers specifically in the use of SSD 
and SSD-related measures where normalization and statistical 
analysis involve the same image set and, instead, advocate a more 
cautious approach
where normalization of the individual subject images
to the statistical reference space occurs through the 
corresponding (statistically independent) anatomical images.

@@

\section{Introduction}
Motivating the findings of a recent paper, the authors related their
puzzlement at the unusually high number of papers reporting 
significant correlations in fMRI studies \cite{Vul2009}. During the 
resulting investigation, the authors discovered the permeating
presence of a particular bias characterized by such related terms as ``circularity''
and ``non-independence''.
While related literature has focused on the presence of such bias in 
fMRI studies \cite{Vul2009,Vul2010} and, more generally, in neuroscience research 
\cite{Kriegeskorte2009}, we show how this can potentially extend to what has
become a fundamental neuroimaging analysis technique---voxel-based analysis.

Similarly, we were perplexed by the disparity in statistical results of one of 
our recent analyses \cite{Stone2011} using the same DTI data and same 
voxel-based analysis technique but different normalization approaches.  Despite 
distinctively better alignments with one approach, there was little 
difference in the statistical results between the two approaches.  
It was eventually discovered that the 
incommensurability between alignment quality and effect size in our
analysis was also a due to a subtle, yet significant circularity bias
which we explain below in the context of FA population studies.  
%Due to the public availability and widespread use of
%these voxel-based analysis tools we show 
%how this source of bias can enter into voxel-based studies focusing on
%population-based FA .

@@

The seminal work of Basser et al. \citep{Basser1994a,Basser1994} established diffusion tensor imaging (DTI) as a viable investigatory MRI technique.  DTI's sensitivity to brain architecture \citep{Basser1996,Assaf2008} enables promising analysis possibilities assessing neuro-structural differences in cross-population studies \citep{Kubicki2005,Arnone2008,Kantarci2010,Rametti2010}.
% including subjects with traumatic brain injury (TBI) \citep{Caeyenberghs2010,Jiang2010,Warner2010}.  
Two popular comprehensive software packages for assessing population differences include the SPM Matlab-based toolkit%
\footnote{
http://www.fil.ion.ucl.ac.uk/spm/
}
for voxel-based morphometry (VBM)
and the tract-based spatial statistics (TBSS) framework \citep{Smith2006}.%
\footnote{
http://www.fmrib.ox.ac.uk/fsl/tbss/index.html
} 
VBM \citep{Ashburner2001} continues to find application to FA studies \citep{Kakeda2010,Takao2010,Preziosa2011} despite concerns about its general validity \citep{Bookstein2001,Davatzikos2004} and with respect to DTI \citep{Jones2005,Chung2008}.  The TBSS framework was partially developed in response to these concerns in which voxel values are projected onto a template white matter skeleton for increased statistical power.  
%(e.g. \citep{Arnone2008,Bodini2009}).

Although there are substantive differences in normalization
between these and other analysis approaches, a core
commonality includes direct FA-to-FA registration using various similarity metrics.  The sum of squared differences (SSD) image similarity metric is perhaps the easiest to interpret (it drives the image intensity difference to zero), is computationally efficient (just compute the image difference and gradient), and is therefore widely used.  Some of the most popular registration methods (e.g. SPM, 
Demons \cite{Thirion1998}, FSL's FNIRT, AIR \citep{Woods1998}, and LDDMM \citep{Beg2005}) rely on this metric and yield reasonable performance levels \citep{Klein2009}.%
%\footnote{
%In addition to TBSS and SPM, other popular packages in the neuroimaging community 
%which use the SSD metric include AIR \citep{Woods1998} and LDDMM \citep{Beg2005}.
%}

However, suppose the SSD metric is used to find a set of $M$ transformations that map a population of $M$ FA images, $\{I_1, I_2, \ldots, I_M\}$, to a representative FA template, $J$, for statistical testing between groups. This results in the transformations,
$\boldsymbol{\mathcal{T}} = \{\mathcal{T}_1, \mathcal{T}_2,\ldots,\mathcal{T}_M \}$, which minimize the SSD metric over the population, i.e.
\begin{align}
  \boldsymbol{\mathcal{T}} = \underset{\mathcal{T}_1, \ldots, \mathcal{T}_M}{\operatorname{argmin}}
    \sum_{m=1}^M\sum_{n=1}^N \left( I^n_m( \mathcal{T}_m(\mathbf{x}) ) - J \right)^2.
\end{align}
Switching the order of the summations (since the transformations are mutually independent) and recognizing that a principal criterion for the selection of $J$ is such that it be a good approximation of the mean of the aligned images, it becomes apparent that the transformation solution,
\begin{align}\label{eq:variance}
  \boldsymbol{\mathcal{T}} = \underset{\mathcal{T}_1, \ldots, \mathcal{T}_M}{\operatorname{argmin}}
    \sum_{n=1}^N \underbrace{\sum_{m=1}^M \left( I^n_m( \mathcal{T}_m(\mathbf{x}) ) - J \right)^2}_{
    \propto\mathrm{\,voxelwise\,variance}},
\end{align}
is that which minimizes the voxelwise group variance.  Reducing the group variance will also directly affect the value of a standard population statistic, the Student's $t$-test.%
\footnote{
Based on this derivation, one would also expect the popular Demons metric to be susceptible to this erroneous increase in statistical power due to identical objective functions although a more aggressive displacement/gradient at each iteration will relatively exacerbate this effect for a given number of iterations.  
Other standard similarity metrics will also exhibit this
trend to some degree as they minimize a local measure of intensity-based distance between the subject and the template.
}

@@

Although various processing choices mitigate the effects of such bias,
such as Gaussian smoothing during and following normalization, 
nonparametric testing \citep{Rorden2007}, 
statistical analysis based on orthogonal projections onto the white matter skeleton 
\cite{Smith2006}, multiple comparisons corrections \cite{Nichols2003}, and the use of more robust similarity metrics, 
such choices are often made ad hoc and/or post hoc and, to our 
knowledge, not with the realization of the potential
for circularity bias described by Eqn. (\ref{eq:variance}).

Instead of direct FA-to-FA template mapping for 
alignment to the normalized space, we advocate a statistically 
independent alignment protocol in which the FA images are mapped to 
the common reference space via intra-subject FA-to-T1 mapping 
followed by T1-to-T1 template alignment thus eliminating all direct
FA-to-FA registrations. The key point is that the intensity value differences used 
to assess alignment (i.e. the similarity metric) in FA-to-T1 and T1-to-T1 
normalizations are statistically independent of the intensity value differences used for 
the post-alignment numerical analysis.  It should be noted that this discussion
is certainly not limited to studies involving FA.  
Such bias can occur anytime the images used for normalization
are non-independent of the statistical analysis as defined by the similarity
metric and statistical testing.

Two sets of experiments establish that the similarity metric
can significantly affect voxel-based analysis outcomes and that,
based on the previous description of circularity bias,
the relationship in performance variability and similarity metric
is to be expected.  Specifically, we show that whereas SSD and Demons
substantially overestimate statistical significance, MI and CC are less susceptible although the
effect is not
negligible.  Furthermore, we also show that this effects persists
even with the common practice of post-normalization smoothing prior
to mass univariate statistical testing.

Given this general overview, the experimental sets can be briefly
characterized as follows:
\begin{itemize}
  \item {\it Simulated DTI Data}:  Using a simulated DTI generator,
  FA images based on male/female populations were generated in situ
  thus guaranteeing anatomical alignment prior to processing.
  The simulated images in each group were then ``registered'' to the FA image
  of the defining DTI 
  template using four common similarity metrics (CC, Demons, MI, and SSD).  
  ROC curves based on voxel-based analysis for each of these scenarios demonstrate that 
  effect size is maximized over anatomical alignment and that
  this effect size varies with similarity metric.  Additional
  experiments establish that this trend persists even for commonly used smoothing
  kernels and the TBSS projection step.
  \item {\it Real TBI Data}:  FA data of TBI survivors and controls were 
  normalized to a template using four common similarity metrics
  (CC, Demons, MI, and SSD).  Additional exploration included normalization via
  the subjects' anatomical T1-weighted images as well as the default registration for TBSS.
  Voxelwise statistical testing was also performed with
  and without smoothing.  Even though ground truth is not available for such
  data, the SSD and Demons metrics showed increased statistical power as evidenced
  by a general decrease in voxelwise $p$-values.
\end{itemize}
All major processing steps were performed using the open source Advanced Normalization Tools 
(ANTs)%
\footnote{
http://www.picsl.upenn.edu/ANTs
}
software repository which should facilitate reproducibility for the interested 
reader.

@@

\subsubsection{NKI/Rockland Data for Generation of the Simulated DTI Gender Cohort}
Data from the first 14 weeks of prospective data-sharing initiative sponsored by the 1000 Functional Connectomes Project (FCP)%
\footnote{
In support of open science, the 1000 Functional Connectomes Project (FCP---http://fcon\_1000.projects.nitrc.org) 
was initiated on December 11, 2009 by various members of the MRI community \citep{Biswal2010}.  Motivated by the absence of neuroimaging data for research, this initiative seeks to form collaborative partnerships with imaging institutions for sharing well-documented multi-modal image sets accompanied by phenotypic data.  Commitments from institutions such as the NYU Institute for Pediatric Neuroscience and the Nathan Kline Institute (NKI)--Rockland have resulted in prospective data repositories
currently distributed to the public via the Neuroimaging Informatics Tools and Resources Clearinghouse (NITRC---http://www.nitrc.org).
}
were downloaded on January 15, 2011 comprised of 74 subjects (average age in years: 32.4 $\pm$ 17.8).  
Due to various issues (e.g. lack of corresponding T1-weighted image, failed DTI reconstruction, and age matching requirements) only 62 subjects (21 females and 41 males).  Each imaging session for each subject produced a 64-directional diffusion tensor imaging scan (parameters: conventional single-shot spin echo
EPI pulse sequence, TR = 10000 ms, TE = 91 ms, axial acquisition, 
voxel size = $2 \times 2 \times 2$ mm$^3$) and a T1-weighted anatomical scan (parameters:  TR = 2500 ms, b = 1000 s/mm$^2$ for each direction, 58 contiguous slices of 2.0 mm thickness,
TI = 1200 ms, TE = 3.5 ms, flip angle = 8$^\circ$, 192 contiguous slices of 
1.0 mm thickness, FOV = 256 $\times$ 256 mm$^2$, and voxel size = 1 mm$^3$).  Images were anonymized including defacing of the T1-weighted images prior to upload. Diffusion tensor data associated with each of the following data sets were reconstructed from the diffusion weighted sequences using Camino%
\footnote{
http://camino.org.uk/
}---an open source toolkit for diffusion MRI processing and 
analysis \citep{Cook2006} in combination with ANTs-based 
registration for motion correction of the DTI sequence.

@@

To provide realistic ground truth FA data, we implemented 
the method of Van Hecke et al. \cite{van-hecke2009} which 
we have also made available to the public
\cite{Tustison2011}.%
\footnote{
http://hdl.handle.net/10380/3315
}  
Given a DTI atlas, a b0 image, a scheme file indicating
the gradient directions, and 
a cohort of aligned DTI data (to model intersubject variability), the
algorithm generates
simulated data sets with specified pathology and noise.
For the simulated experiments described in the next section we
created a T1-weighted population-specific template from 30 randomly chosen T1-weighted images of
the NKI/Rockland data set previously described.  
The resulting
template was then registered to the MNI152 template.  
The remaining 32 T1-weighted images were also registered to the population-specific 
template resulting in a composite transform for each subject describing
the mapping from the individual subject's T1-weighted image domain to 
the MNI152 template.

To obtain the mappings from the DTI space to the population-specific
template, the following steps were taken:
\begin{itemize}
  \item For each anatomical T1-weighted image, segment the white matter (in ANTs, one can use N4 bias correction \citep{Tustison2010} and Atropos $n$-tissue segmentation \citep{Avants2011a}).
  \item Rigidly transform each subject's average DWI to the bias-corrected anatomical image ($\mathcal{R}^i_{dwi}$).
  \item Perform distortion correction on the FA to anatomical mapping 
        by localized nonrigid registration
        of rigidly transformed FA map ($=\mathcal{R}^i_{dwi} \mathrm{FA}$) using
        the white matter mask obtained in a previous step ($\mathcal{T}^i_{FA}$).
  \item For each subject, concatenate the transforms which maps each
        subject's FA to the template ($\mathcal{T}^i_{total} = \mathcal{T}^i_{template} \circ 
        \mathcal{T}^i_{FA} \circ \mathcal{R}^i_{dwi} \mathrm{FA}$).
  \item Warp each subject's FA map to the template using $\mathcal{T}^i_{total}$.  
%  Equivalently, one could warp the mean diffusion (MD), axial diffusion (AD) or any 
%  other scalar derivable from DTI to the template.  
\end{itemize}
DTI 
were reoriented using the PPD method \cite{alexander2001}.  The 62
aligned DTI were averaged using the log-euclidean framework 
\cite{arsigny2006} to create the DTI atlas.  Similarly, the 62 aligned
b0 images were averaged to create the representative b0 image
for simulated DTI creation.  The derived FA image from the DTI atlas and 
the b0 image are shown in 
Fig. \ref{fig:simulated}.  An FA mask was created by thresholding 
the FA image of the DTI atlas at 0.2.

The 41 aligned male DTI were used to generate 18 ``Control'' 30-direction
diffusion-weighted images
with no simulated noise nor pathology.  Similarly, the 21 aligned female
DTI were used to generate 19 ``Experimental'' diffusion-weighted image sets.
DTI were reconstructed using Camino which were then used to create
individual FA images.
The number of subjects in each
group was chosen based on the median number of control and experimental
subjects from the sampling of FA studies described in 
\cite{van-hecke2009}.

@@

@@

\begin{figure*}
\begin{center}
\begin{tabular}{ccc}
  \includegraphics[width=57mm]{ISV_SpeSenAlreadyAligned.pdf} &
  \includegraphics[width=57mm]{ISV_SpeSenAlreadyAlignedSmooth0.pdf} &
  \includegraphics[width=57mm]{ISV_SpeSenAlreadyAlignedSmooth1.pdf} \\
  \includegraphics[width=57mm]{ISV_CDFAlreadyAligned.pdf} &
  \includegraphics[width=57mm]{ISV_CDFAlreadyAlignedSmooth0.pdf} &
  \includegraphics[width=57mm]{ISV_CDFAlreadyAlignedSmooth1.pdf} \\
\end{tabular}
\caption{Post-normalization smoothing analysis of the simulated data for
fwhm kernel sizes of 0 mm, 4 mm, and 8 mm.  Top row:  
ROC curves plotting $(1 - specificity)$ vs. $sensitivity$ for the
different similarity metrics for voxelwise analysis.  Smoothing
tends to ameliorate the bias effect but the relative strength
continues to correlate with similarity metric.  Bottom row:
A relative 
increase in (false) statistical power induced by SSD and Demons, as 
described in the text, resulted in more voxels having lower 
$p$-values.  This will be manifested as a left-shifted CDF 
curve which is apparent in the SSD and Demons metrics for all selected smoothing kernel sizes.
}
\label{fig:simulated_plots_smooth}
\end{center}        
\end{figure*}

\begin{figure}
\begin{center}
\begin{tabular}{cc}
  \includegraphics[width=41mm]{smooth0_truth.png} &
  \includegraphics[width=41mm]{smooth0_demons.png} \\
  (a) & (b)
\end{tabular}
\caption{(a) Mid-axial slice of the DTI atlas-derived FA map where the 
significant differences ($p \leq 0.05$, smoothing = 4 mm fwhm) between control (male) and 
experimental (female) groups are labeled in red.  (b) Same mid-axial 
slice showing an increase in significant regions following ANTs 
registration employing the Demons metric.
}
\label{fig:simulated_projection}
\end{center}        
\end{figure}

>>>>

  \usepackage{lineno}
  \usepackage{makecell}
\usepackage{amssymb}
\usepackage{color}
\usepackage{multirow,booktabs,ctable,array}
\usepackage{lscape}
\usepackage{amsmath}
\usepackage{ulem}
\usepackage{setspace}
\usepackage{listings}
\usepackage{float}

@@

%% The amssymb package provides various useful mathematical symbols
%% The amsthm package provides extended theorem environments
% \usepackage{amsthm}

@@

\begin{abstract}
Recent discussions within the neuroimaging community have highlighted the
problematic presence of selection bias in experimental design.  Although 
initially centering around the selection of voxels during the course of
fMRI studies, we demonstrate how this bias can 
potentially corrupt voxel-based analysis with a particular emphasis on
population studies of fractional anisotropy (FA).  
For such studies, template-based registration plays a critical role
in which a representative template serves as the normalized space for 
group alignment.  A standard approach maps each subject's FA image
to a FA template before performing statistical comparisons between different groups.
We illustrate how this analysis strategy can induce selection
bias which varies in strength with the similarity metric used in
registration.  Specifically, we analytically demonstrate that the popular sum of squared
difference (SSD) intensity metric explicitly maximizes effect size rather than
anatomical alignment---an experimental design flaw referred to as `circularity bias.'  
Through the use of both simulated diffusion data and real data involving
traumatic brain injury survivors, we explore the magnitude of this bias
and how it varies with similarity metric under the hypothesis that while
SSD-related metrics, such as Demons, will manifest similar effects, 
other metrics which 
are not formulated based on absolute intensity differences 
(e.g. mutual information (MI) and cross correlation (CC)) will produce
less of an effect. 
Consequently, given the variability in voxel-based analysis outcomes 
with similarity metric choice,
we caution researchers specifically in the use of SSD 
and SSD-related measures where normalization and statistical 
analysis involve the same image set and, instead, advocate a more 
cautious approach
where normalization of the individual subject images
to the statistical reference space occurs through the 
corresponding (statistically independent) anatomical images.  A second
alternative is to use similarity terms that are less sensitive to this bias.

@@

\section{Introduction}
Motivating the findings of a recent paper, the authors related their
puzzlement at the unusually high number of papers reporting 
significant correlations in fMRI studies \citep{Vul2009}. During the 
resulting investigation, the authors discovered the permeating
presence of a particular bias characterized by such related terms as ``circularity''
and ``non-independence''.
While related literature has focused on the presence of such bias in 
fMRI studies \citep{Vul2009,Vul2010} and, more generally, in neuroscience research 
\citep{Kriegeskorte2009}, we show how this can potentially extend to what has
become a fundamental neuroimaging analysis technique---voxel-based analysis.

Similarly, we were perplexed by the disparity in statistical results of one of 
our recent analyses \citep{Stone2011} using the same DTI data and same 
voxel-based analysis technique but different normalization approaches.  Despite 
distinctively better alignments with one approach, there was little 
difference in the statistical results between the two approaches.  
It was eventually discovered that the 
incommensurability between alignment quality and effect size in our
analysis was also a due to a subtle, yet significant circularity bias
which we explain below in the context of FA population studies.  
%Due to the public availability and widespread use of
%these voxel-based analysis tools we show 
%how this source of bias can enter into voxel-based studies focusing on
%population-based FA .

@@

The seminal work of Basser et al. \citep{Basser1994a,Basser1994} established diffusion tensor imaging (DTI) as a viable investigatory MRI technique.  DTI's sensitivity to brain architecture \citep{Basser1996,Assaf2008} enables promising analysis possibilities assessing neuro-structural differences in cross-population studies \citep{Kubicki2005,Arnone2008,Kantarci2010,Rametti2010}.
% including subjects with traumatic brain injury (TBI) \citep{Caeyenberghs2010,Jiang2010,Warner2010}.  
Two popular comprehensive software packages for assessing population differences include the SPM Matlab-based toolkit%
\footnote{
http://www.fil.ion.ucl.ac.uk/spm/
}
for voxel-based morphometry (VBM)
and the tract-based spatial statistics (TBSS) framework \citep{Smith2006}.%
\footnote{
http://www.fmrib.ox.ac.uk/fsl/tbss/index.html
} 
VBM \citep{Ashburner2001} continues to find application to FA studies
\citep{Kakeda2010,Takao2010,Preziosa2011} despite concerns about its
general validity \citep{Bookstein2001,Davatzikos2004} in particular with respect to DTI \citep{Jones2005,Chung2008}.  The TBSS framework was partially developed in response to these concerns in which voxel values are projected onto a template white matter skeleton for increased statistical power.  
%(e.g. \citep{Arnone2008,Bodini2009}).

Although there are substantive differences in normalization
between these and other analysis approaches, a core
commonality includes direct FA-to-FA registration using various similarity metrics.  The sum of squared differences (SSD) image similarity metric is perhaps the easiest to interpret (it drives the image intensity difference to zero), is computationally efficient (compute the image difference and gradient), and is therefore widely used.  Some of the most popular registration methods (e.g. SPM, 
Demons \cite{Thirion1998}, FSL's FNIRT and LDDMM \citep{Beg2005}) rely on this metric and yield reasonable performance levels \citep{Klein2009}.%
%\footnote{
%In addition to TBSS and SPM, other popular packages in the neuroimaging community 
%which use the SSD metric include AIR \citep{Woods1998} and LDDMM \citep{Beg2005}.
%}

However, suppose the SSD metric is used to find a set of $M$ transformations that map a population of $M$ FA images, $\{I_1, I_2, \ldots, I_M\}$, to a representative FA template, $J$, for statistical testing between groups. This results in the transformations,
$\boldsymbol{\mathcal{T}} = \{\mathcal{T}_1, \mathcal{T}_2,\ldots,\mathcal{T}_M \}$, which minimize the SSD metric over the population, i.e.
\begin{align}
  \boldsymbol{\mathcal{T}} = \underset{\mathcal{T}_1, \ldots, \mathcal{T}_M}{\operatorname{argmin}}
    \sum_{m=1}^M\sum_{n=1}^N \left( I^n_m( \mathcal{T}_m(\mathbf{x}) ) - J \right)^2.
\end{align}
Switching the order of the summations and recognizing that a principal criterion for the selection of $J$ is such that it be a good approximation of the mean of the aligned images, it becomes apparent that the transformation solution,
\begin{align}\label{eq:variance}
  \boldsymbol{\mathcal{T}} = \underset{\mathcal{T}_1, \ldots, \mathcal{T}_M}{\operatorname{argmin}}
    \sum_{n=1}^N \underbrace{\sum_{m=1}^M \left( I^n_m( \mathcal{T}_m(\mathbf{x}) ) - J \right)^2}_{
    \propto\mathrm{\,voxelwise\,variance}},
\end{align}
is that which minimizes the voxelwise group variance.  Reducing the group variance will also directly affect the value of a standard population statistic, the Student's $t$-test.%
\footnote{
Based on this derivation, one would also expect the popular Demons metric to be susceptible to this erroneous increase in statistical power due to identical objective functions although a more aggressive displacement/gradient at each iteration will relatively exacerbate this effect for a given number of iterations.  
Other standard similarity metrics will also exhibit this
trend to some degree as they minimize a local measure of intensity-based distance between the subject and the template.
}

@@

Various processing choices may mitigate the effects of such bias.  Gaussian smoothing during and following normalization, 
nonparametric testing \citep{Rorden2007}, 
statistical analysis based on orthogonal projections onto the white matter skeleton 
\citep{Smith2006}, multiple comparisons corrections \citep{Nichols2003},
and the use of more robust similarity metrics may all impact the
outcome.  However, such choices are often made ad hoc and/or post hoc and, to our 
knowledge, not with the realization of the potential
for circularity bias described by Eqn. (\ref{eq:variance}).

Instead of direct FA-to-FA template mapping for 
alignment to the normalized space, we advocate a statistically 
independent alignment protocol in which the FA images are mapped to 
the common reference space via intra-subject FA-to-T1 mapping 
followed by T1-to-T1 template alignment thus eliminating all direct
FA-to-FA registrations. The key point is that the intensity value differences used 
to assess alignment (i.e. the similarity metric) in FA-to-T1 and T1-to-T1 
normalizations are statistically independent of the intensity value differences used for 
the post-alignment numerical analysis.  It should be noted that this discussion
is certainly not limited to studies involving FA.  
Such bias can occur anytime the images used for normalization
are non-independent of the statistical analysis as defined by the similarity
metric and statistical testing. {\color{red}{ For example, following Eqn. (\ref{eq:variance}),
circularity could theoretically be an issue for VBM if one segmented 
the template image to obtain the gray matter density (GMD)
template image and then registered individual subject GMD
images to the GMD template image using SSD-related similarity measures 
(e.g. \cite{Good2001}, FSL-VBM,%
\footnote{
{\color{red}{http://www.fmrib.ox.ac.uk/fsl/fslvbm/index.html}}
}%
and DARTEL VBM%
\footnote{
{\color{red}{http://www.fil.ion.ucl.ac.uk/~john/misc/VBMclass10.pdf}}
}%
).}}

Two sets of experiments establish that the similarity metric
can significantly affect voxel-based analysis outcomes and that,
based on the previous description of circularity bias,
the relationship in performance variability and similarity metric
is to be expected.  Specifically, we show that whereas SSD and Demons
substantially overestimate statistical significance, MI and CC are less susceptible although the
effect is not
negligible.  Furthermore, we also show that this effect persists
even with the common practice of post-normalization smoothing prior
to mass univariate statistical testing.

Given this general overview, the experimental sets can be briefly
characterized as follows:
\begin{itemize}
  \item {\it Simulated DTI Data}:  Using a simulated DTI generator,
  FA images based on male/female populations were generated in situ
  thus guaranteeing anatomical alignment prior to processing.
  The simulated images in each group were then ``registered'' to the FA image
  of the defining DTI 
  template using four common similarity metrics (CC, Demons, MI, and SSD).  
  ROC curves based on voxel-based analysis for each of these scenarios demonstrate that 
  effect size is maximized over anatomical alignment and that
  this effect size varies with similarity metric.  Additional
  experiments establish that this trend persists even for commonly used smoothing
  kernels and the TBSS projection step.
  \item {\it Real TBI Data}:  FA data of TBI survivors and controls were 
  normalized to a template using four common similarity metrics
  (CC, Demons, MI, and SSD).  Additional exploration included normalization via
  the subjects' anatomical T1-weighted images as well as the default registration for TBSS.
  Voxelwise statistical testing was also performed with
  and without smoothing.  Even though ground truth is not available for such
  data, the SSD and Demons metrics showed increased statistical power as evidenced
  by a general decrease in voxelwise $p$-values.
\end{itemize}
All major processing steps were performed using publicly available software including the open source Advanced Normalization Tools 
(ANTs)%
\footnote{
http://www.picsl.upenn.edu/ANTs
}
which should facilitate reproducibility for the interested 
reader.

@@

\subsubsection{NKI/Rockland Data for Generation of the Simulated DTI Gender Cohort}
Data from the first 14 weeks of prospective data-sharing initiative sponsored by the 1000 Functional Connectomes Project (FCP)%
\footnote{
In support of open science, the 1000 Functional Connectomes Project (FCP---http://fcon\_1000.projects.nitrc.org) 
was initiated on December 11, 2009 by various members of the MRI community \citep{Biswal2010}.  Motivated by the absence of neuroimaging data for research, this initiative seeks to form collaborative partnerships with imaging institutions for sharing well-documented multi-modal image sets accompanied by phenotypic data.  Commitments from institutions such as the NYU Institute for Pediatric Neuroscience and the Nathan Kline Institute (NKI)--Rockland have resulted in prospective data repositories
currently distributed to the public via the Neuroimaging Informatics Tools and Resources Clearinghouse (NITRC---http://www.nitrc.org).
}
were downloaded on January 15, 2011 comprised of 74 subjects (average age in years: 32.4 $\pm$ 17.8).  
Due to various issues (e.g. lack of corresponding T1-weighted image,
failed DTI reconstruction, and age matching requirements) only 62
subjects (21 females and 41 males) were available.  Each imaging session for each subject produced a 64-directional diffusion tensor imaging scan (parameters: conventional single-shot spin echo
EPI pulse sequence, TR = 10000 ms, TE = 91 ms, axial acquisition, 
voxel size = $2 \times 2 \times 2$ mm$^3$) and a T1-weighted anatomical scan (parameters:  TR = 2500 ms, b = 1000 s/mm$^2$ for each direction, 58 contiguous slices of 2.0 mm thickness,
TI = 1200 ms, TE = 3.5 ms, flip angle = 8$^\circ$, 192 contiguous slices of 
1.0 mm thickness, FOV = 256 $\times$ 256 mm$^2$, and voxel size = 1 mm$^3$).  Images were anonymized including defacing of the T1-weighted images prior to upload. Diffusion tensor data associated with each of the following data sets were reconstructed from the diffusion weighted sequences using Camino%
\footnote{
http://camino.org.uk/
}---an open source toolkit for diffusion MRI processing and 
analysis \citep{Cook2006} in combination with ANTs-based 
registration for motion correction of the DTI sequence.

@@

To provide realistic ground truth FA data, we implemented 
the method of Van Hecke et al. \citep{van-hecke2009} which 
we have also made available to the public
\citep{Tustison2011}.%
\footnote{
http://hdl.handle.net/10380/3315
}  
Given a DTI atlas, a b0 image, a scheme file indicating
the gradient directions, and 
a cohort of aligned DTI data (to model intersubject variability), the
algorithm generates
simulated data sets with specified pathology and noise.
For the simulated experiments described in the next section we
created a T1-weighted population-specific template from 30 randomly chosen T1-weighted images of
the NKI/Rockland data set previously described.  
The resulting
template was then registered to the MNI152 template.  
The remaining 32 T1-weighted images were also registered to the population-specific 
template resulting in a composite transform for each subject describing
the mapping from the individual subject's T1-weighted image domain to 
the MNI152 template.

To obtain the mappings from the DTI space to the population-specific
template, the following steps were taken:
\begin{itemize}
  \item For each anatomical T1-weighted image, segment the white matter (in ANTs, one can use N4 bias correction \citep{Tustison2010} and Atropos $n$-tissue segmentation \citep{Avants2011a}).
  \item Rigidly transform each subject's average DWI to the bias-corrected anatomical image ($\mathcal{R}^i_{dwi}$).
  \item Perform distortion correction on the {\color{red}{DTI}} to anatomical mapping 
        by localized nonrigid registration
        of rigidly transformed FA map ($=\mathcal{R}^i_{dwi} \mathrm{FA}$) using
        the white matter mask obtained in a previous step ($\mathcal{T}^i_{FA}$).
  \item For each subject, concatenate the transforms which maps each
        subject's {\color{red}{DTI}} to the template ($\mathcal{T}^i_{total} = \mathcal{T}^i_{template} \circ 
        \mathcal{T}^i_{FA} \circ \mathcal{R}^i_{dwi} \mathrm{FA}$).
  \item Warp each subject's {\color{red}{DTI}} map to the template using $\mathcal{T}^i_{total}$.  
%  Equivalently, one could warp the mean diffusion (MD), axial diffusion (AD) or any 
%  other scalar derivable from DTI to the template.  
\end{itemize}
DTI 
were reoriented using the PPD method \cite{alexander2001}.  The 62
aligned DTI were averaged using the log-euclidean framework 
\cite{arsigny2006} to create the DTI atlas.  Similarly, the 62 aligned
b0 images were averaged to create the representative b0 image
for simulated DTI creation.  The derived FA image from the DTI atlas and 
the b0 image are shown in 
Fig. \ref{fig:simulated}.  An FA mask was created by thresholding 
the FA image of the DTI atlas at 0.2.

The 41 aligned male DTI were used to generate 18 ``Control'' 30-direction
diffusion-weighted images
with no simulated noise nor pathology.  Similarly, the 21 aligned female
DTI were used to generate 19 ``Experimental'' diffusion-weighted image sets.
These simulated DTI were reconstructed using Camino which yielded
individual FA images.
The number of subjects in each
group was chosen based on the median number of control and experimental
subjects from the sampling of FA studies described in 
\cite{van-hecke2009}.

@@

\subsection{Statistical methods} 
We employ two complementary tools for examining the impact of image 
registration strategy on statistical outcome.  The receiver operating
characteristic (ROC) and the cumulative distribution function (CDF).
We provide a brief summary of these tools and their purpose in this study.
\subsubsection{ROC analysis} 
The ROC may be used to look at the performance of a continuous model
against a known classification when the threshold on the model is
varied.  In this study, the ground-truth is determined by the voxels
that---via simulation---we know have a significant difference between
groups.  We choose $p<0.05$ to determine the
ground truth.  This leads to two classes of voxels: those that are
significantly different between groups and those that are not.  As we
vary the threshold on the statistical map that is generated by a
registration algorithm, we can compare the identified voxels against
this ground truth and draw the ROC \citep{Fitzpatrick1998,Wenzel2010}.

\subsubsection{Cumulative distribution functions} 
Empirically constructed CDFs have been used with maps of $p$-values,
derived from registration, to visually quantify detection power
\citep{Brun2008,Hua2009}.  In this case, one examines the area under the
CDF that is to the left of some statistical cutoff, usually $p<0.05$.  If one assumes
that the method is not producing biased results, then the method that
has a greater area under the CDF, left of cutoff, has greater
detection power.

@@

%\begin{landscape}
\begin{figure*}
\begin{center}
\begin{tabular}{ccc}
  \includegraphics[width=57mm]{ISV_SpeSenAlreadyAligned.pdf} &
  \includegraphics[width=57mm]{ISV_SpeSenAlreadyAlignedSmooth0.pdf} &
  \includegraphics[width=57mm]{ISV_SpeSenAlreadyAlignedSmooth1.pdf} \\
  \includegraphics[width=57mm]{ISV_CDFAlreadyAligned.pdf} &
  \includegraphics[width=57mm]{ISV_CDFAlreadyAlignedSmooth0.pdf} &
  \includegraphics[width=57mm]{ISV_CDFAlreadyAlignedSmooth1.pdf} \\
\end{tabular}
\caption{Post-normalization smoothing analysis of the simulated data for
fwhm kernel sizes of 0 mm, 4 mm, and 8 mm.  The ground truth data is
smoothed in the same way as the normalized data.  Top row:  
ROC curves plotting $(1 - specificity)$ vs. $sensitivity$ for the
different similarity metrics for voxelwise analysis.  Smoothing
tends to ameliorate the bias effect but the relative strength
continues to correlate with similarity metric.  Bottom row:
A relative 
increase in (false) statistical power induced by SSD and Demons, as 
described in the text, resulted in more voxels having lower 
$p$-values.  This will be manifested as a left-shifted CDF 
curve which is apparent in the SSD and Demons metrics for all selected
smoothing kernel sizes.  In summary, improvements in detection power
(if judged by the area under the CDF that is below 0.05) do not imply results that are closer to the
truth (which is shown by examining total area under the ROC curve).
}
\label{fig:simulated_plots_smooth}
\end{center}        
\end{figure*}
%\end{landscape}
